---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gpkg - Utilities for OGC GeoPackages
<!-- badges: start -->
[![R-CMD-check](https://github.com/brownag/gpkg/actions/workflows/R-CMD-check.yml/badge.svg?branch=main)](https://github.com/brownag/gpkg/actions/workflows/R-CMD-check.yml)
[![gpkg HTML Manual](https://img.shields.io/badge/docs-HTML-informational)](http://humus.rocks/gpkg/)
<!-- badges: end -->

{gpkg} provides high-level wrapper functions to build GeoPackages containing a variety of different data. Reading and writing of spatial ([vector](http://www.gdal.org/drv_geopackage.html) and [gridded](http://www.gdal.org/drv_geopackage_raster.html)) data are done via standard [GDAL](http://www.gdal.org/) utilities provided primarily by the {terra} package. Additional functions are provided to manipulate attributes and tabular data via [RSQLite](https://cran.r-project.org/web/packages/RSQLite/index.html).

<a href="https://raw.githubusercontent.com/brownag/gpkg/main/man/figures/gpkg_sticker_v1.png">
<img src = "https://raw.githubusercontent.com/brownag/gpkg/main/man/figures/gpkg_sticker_v1.png" alt = "gpkg hexsticker" title = "gpkg hexsticker: {gpkg} provides high-level wrapper functions to build GeoPackages containing a variety of different data." width = "35%" height = "35%" hspace="25" vspace="25" align="right"/></a>

## Installation

The package can be installed from GitHub with {remotes}

```r
if (!requireNamespace("remotes")) 
  install.packages("remotes")
remotes::install_github("brownag/gpkg")
```

## Background
### What is a GeoPackage?

[GeoPackage](https://www.geopackage.org/) is an open, standards-based, platform-independent, portable, self-describing, compact format for transferring geospatial information. The [GeoPackage Encoding Standard](https://www.ogc.org/standards/geopackage) describes a set of conventions for storing the following within an SQLite database:

  *  vector features
  
  *  tile matrix sets of imagery and raster maps at various scales
  
  *  attributes (non-spatial data)
  
  *  extensions

## Create a Geopackage

`gpkg_write()` can handle a variety of different input types. Here we start by adding two DEM (GeoTIFF) files.

```{r}
library(gpkg)
library(terra)

dem <- system.file("extdata", "dem.tif", package = "gpkg")
stopifnot(nchar(dem) > 0)
gpkg_tmp <- tempfile(fileext = ".gpkg")

if (file.exists(gpkg_tmp))
  file.remove(gpkg_tmp)

# write a gpkg with two DEMs in it
gpkg_write(
  dem,
  destfile = gpkg_tmp,
  RASTER_TABLE = "DEM1",
  FIELD_NAME = "Elevation"
)
gpkg_write(
  dem,
  destfile = gpkg_tmp,
  append = TRUE,
  RASTER_TABLE = "DEM2",
  FIELD_NAME = "Elevation"
)
```
## Insert Vector Layers

We can also write vector data to GeoPackage. Here we use `terra::writeVector(..., insert = TRUE)` to add a bounding box polygon layer derived from `"DEM1"`.

```{r}
# add bounding polygon vector layer
# TODO: wrap this into gpkg_write() and add support for in-memory rasters
terra::writeVector(
  terra::set.crs(terra::as.polygons(terra::ext(
    gpkg_tables(geopackage(gpkg_tmp))[['DEM1']])
  ), "OGC:CRS84"),
  filename = gpkg_tmp,
  insert = TRUE
)
```

## Read a GeoPackage

`geopackage()` is a constructor that can create a simple container for working with geopackages from several types of inputs. Often you will have a _character_ file path to a GeoPackage (.gpkg) file. Other times you may have a list of tables and layers you want to be in a GeoPackage that does not exist yet. Or, you may have a connection to a GeoPackage database already opened that you want to use. In any case (`character`, `list`, `SQLiteConnection`) there is an S3 method to facilitate creating the basic `geopackage` class provided by {gpkg}.

```{r}
g <- geopackage(gpkg_tmp, connect = TRUE)
g
class(g)
```

## Inspect Contents of GeoPackage

We can list the table names in a GeoPackage with `gpkg_list_tables()` and fetch pointers (SpatRaster, SpatVectorProxy, and lazy data.frame) to the data in them with `gpkg_table()`. We can check the status of the internal `geopackage` class `SQLiteConnection` with `gpkg_is_connected()` and disconnect it with `gpkg_disconnect()`.

```{r}
# enumerate tables
gpkg_list_tables(g)

# inspect tables
gpkg_tables(g)

# still connected
gpkg_is_connected(g)

# disconnect geopackage
gpkg_disconnect(g)
```
