---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gpkg - Utilities for the OGC 'GeoPackage' Format

<!-- badges: start -->
[![R-CMD-check](https://github.com/brownag/gpkg/actions/workflows/R-CMD-check.yml/badge.svg?branch=main)](https://github.com/brownag/gpkg/actions/workflows/R-CMD-check.yml)
[![gpkg HTML Manual](https://img.shields.io/badge/docs-HTML-informational)](http://humus.rocks/gpkg/)
[![CRAN status](https://www.r-pkg.org/badges/version/gpkg)](https://CRAN.R-project.org/package=gpkg)
[![Codecov test coverage](https://codecov.io/gh/brownag/gpkg/branch/main/graph/badge.svg)](https://app.codecov.io/gh/brownag/gpkg?branch=main)
<!-- badges: end -->

High-level wrapper functions to build [Open Geospatial Consortium (OGC) 'GeoPackage' files](https://geopackage.org). [GDAL](http://www.gdal.org/) utilities for read and write of spatial data ([vector](http://www.gdal.org/drv_geopackage.html) and [gridded](http://www.gdal.org/drv_geopackage_raster.html)) are provided via the {[terra](https://cran.r-project.org/package=terra)} package. Additional 'GeoPackage' and 'SQLite' specific functions manipulate attributes and tabular data via the {[RSQLite](https://cran.r-project.org/package=RSQLite)} package.


<a href="https://raw.githubusercontent.com/brownag/gpkg/main/man/figures/gpkg_sticker_v1.png">
<img src = "https://raw.githubusercontent.com/brownag/gpkg/main/man/figures/gpkg_sticker_v1.png" alt = "gpkg hexsticker" title = "gpkg hexsticker: {gpkg} provides high-level wrapper functions to build GeoPackages containing a variety of different data." width = "35%" height = "35%" hspace="25" vspace="25" align="right"/></a>

## Installation

The package can be installed from GitHub with {remotes}

```r
if (!requireNamespace("remotes")) 
  install.packages("remotes")
remotes::install_github("brownag/gpkg")
```

## Background
### What is a GeoPackage?

[GeoPackage](https://www.geopackage.org/) is an open, standards-based, platform-independent, portable, self-describing, compact format for transferring geospatial information. The [GeoPackage Encoding Standard](https://www.ogc.org/standards/geopackage) describes a set of conventions for storing the following within an SQLite database:

  *  vector features
  
  *  tile matrix sets of imagery and raster maps at various scales
  
  *  attributes (non-spatial data)
  
  *  extensions

## Create a Geopackage

`gpkg_write()` can handle a variety of different input types. Here we start by adding two DEM (GeoTIFF) files.

```{r}
library(gpkg)
library(terra)

dem <- system.file("extdata", "dem.tif", package = "gpkg")
stopifnot(nchar(dem) > 0)
gpkg_tmp <- tempfile(fileext = ".gpkg")

if (file.exists(gpkg_tmp))
  file.remove(gpkg_tmp)

# write a gpkg with two DEMs in it
gpkg_write(
  dem,
  destfile = gpkg_tmp,
  RASTER_TABLE = "DEM1",
  FIELD_NAME = "Elevation"
)

gpkg_write(
  dem,
  destfile = gpkg_tmp,
  append = TRUE,
  RASTER_TABLE = "DEM2",
  FIELD_NAME = "Elevation",
  NoData = -9999
)
```

## Insert Vector Layers

We can also write vector data to GeoPackage. Here we use `gpkg_write()` to add a bounding box polygon layer derived from extent of `"DEM1"`.

```{r}
# add bounding polygon vector layer via named list
r <- gpkg_tables(geopackage(gpkg_tmp))[['DEM1']]
v <- terra::as.polygons(r, ext = TRUE)
gpkg_write(list(bbox = v), destfile = gpkg_tmp, append = TRUE)
```

## Insert Attribute Table

Similarly, `data.frame`-like objects (non-spatial "attributes") can be written to GeoPackage.

```{r}
z <- data.frame(a = 1:10, b = LETTERS[1:10])
gpkg_write(list(myattr = z), destfile = gpkg_tmp, append = TRUE)
```

## Read a GeoPackage

`geopackage()` is a constructor that can create a simple container for working with geopackages from several types of inputs. Often you will have a _character_ file path to a GeoPackage (.gpkg) file. Other times you may have a list of tables and layers you want to be in a GeoPackage that does not exist yet. Or, you may have a connection to a GeoPackage database already opened that you want to use. In any case (`character`, `list`, `SQLiteConnection`) there is an S3 method to facilitate creating the basic `geopackage` class provided by {gpkg}.

```{r}
g <- geopackage(gpkg_tmp, connect = TRUE)
g
class(g)
```

## Inspect Contents of GeoPackage

We can list the table names in a GeoPackage with `gpkg_list_tables()` and fetch pointers (SpatRaster, SpatVectorProxy, and lazy data.frame) to the data in them with `gpkg_table()`. We can check the status of the internal `geopackage` class `SQLiteConnection` with `gpkg_is_connected()` and disconnect it with `gpkg_disconnect()`.

```{r}
# enumerate tables
gpkg_list_tables(g)

# inspect tables
gpkg_tables(g)
```

### Lazy Tables and {dplyr} Integration

Two methods for 'lazy' access of table contents are available:


#### Method 1: `lazy.frame()`

`lazy.frame()` is a low-frills `data.frame` result containing important table information, but not values.  The `PRAGMA table_info()` is stored as a nested data.frame `table_info`. This representation has no dependencies beyond {RSQLite} and is efficient for inspection of table structure and attributes. Though it is less useful for data analysis.

```{r}
head(lazy.frame(g))
```

#### Method 2: `dplyr.frame()`

With the `dplyr.frame()` method you access a specific table (by name) and get a "lazy" `tibble` object referencing that table. This is achieved via {dplyr} and the {dbplyr} database connection to the GeoPackage via the {RSQLite} driver. The resulting object's data can be used in more complex analyses by using other {dbplyr}/{tidyverse} functions.

For example, we inspect the contents of the `gpkg_contents` table that contains critical information on the data contained in a GeoPackage.

```{r}
dplyr.frame(g, "gpkg_contents")
```

As a more complicated example we access the `gpkg_2d_gridded_tile_ancillary` table, and perform some data processing.

We `dplyr::select()` `mean` and `std_dev` columns from the `dplyr::filter()`ed rows where `tpudt_name == "DEM2"`. Finally we materialize a `tibble`  with `dplyr::collect()`:

```{r}
library(dplyr, warn.conflicts = FALSE)

dplyr.frame(g, "gpkg_2d_gridded_tile_ancillary") %>% 
  filter(tpudt_name == "DEM2") %>% 
  select(mean, std_dev) %>% 
  collect()
```

### Managing Connections

Several helper methods are available for checking GeoPackage `SQLiteConnection` status, as well as connecting and disconnecting an existing `geopackage` object (`g`).

```{r}
# still connected
gpkg_is_connected(g)

# disconnect geopackage
gpkg_disconnect(g)

# reconnect
g <- gpkg_connect(g)

# disconnect
gpkg_disconnect(g)
```
