---
title: "Introduction to gpkg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gpkg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("terra", quietly = TRUE) && 
    requireNamespace("dbplyr", quietly = TRUE)
)
```

## Background

`gpkg` provides high-level wrapper functions to build [Open Geospatial Consortium (OGC) 'GeoPackage' files](https://www.geopackage.org/). [GDAL](https://gdal.org/) utilities for read and write of spatial data ([vector](https://gdal.org/en/stable/drivers/vector/gpkg.html) and [gridded](https://gdal.org/en/stable/drivers/raster/gpkg.html)) are provided via the {[terra](https://cran.r-project.org/package=terra)} package. Additional 'GeoPackage' and 'SQLite' specific functions manipulate attributes and tabular data via the {[RSQLite](https://cran.r-project.org/package=RSQLite)} package.

### What is a GeoPackage?

[GeoPackage](https://www.geopackage.org/) is an open, standards-based, platform-independent, portable, self-describing, compact format for transferring geospatial information. The [GeoPackage Encoding Standard](https://www.ogc.org/publications/standard/geopackage/)) describes a set of conventions for storing the following within an SQLite database:

  *  vector features
  
  *  tile matrix sets of imagery and raster maps at various scales
  
  *  attributes (non-spatial data)
  
  *  extensions

## Create a Geopackage

`gpkg_write()` can handle a variety of different input types. Here we start by adding two DEM (GeoTIFF) files.

```{r}
library(gpkg)
library(terra)

dem <- system.file("extdata", "dem.tif", package = "gpkg")
stopifnot(nchar(dem) > 0)
gpkg_tmp <- tempfile(fileext = ".gpkg")

if (file.exists(gpkg_tmp))
  file.remove(gpkg_tmp)

# write a gpkg with two DEMs in it
gpkg_write(
  dem,
  destfile = gpkg_tmp,
  RASTER_TABLE = "DEM1",
  FIELD_NAME = "Elevation"
)

gpkg_write(
  dem,
  destfile = gpkg_tmp,
  append = TRUE,
  RASTER_TABLE = "DEM2",
  FIELD_NAME = "Elevation",
  NoData = -9999
)
```

## Insert Vector Layers

We can also write vector data to GeoPackage. Here we use `gpkg_write()` to add a bounding box polygon layer derived from extent of `"DEM1"`.

```{r}
# add bounding polygon vector layer via named list
r <- gpkg_tables(gpkg_tmp)[['DEM1']]
v <- terra::as.polygons(r, ext = TRUE)
gpkg_write(list(bbox = v), destfile = gpkg_tmp)
```

## Insert Attribute Table

Similarly, `data.frame`-like objects (non-spatial "attributes") can be written to GeoPackage.

```{r}
z <- data.frame(a = 1:10, b = LETTERS[1:10])
gpkg_write(list(myattr = z), destfile = gpkg_tmp)
```

## Read a GeoPackage

`geopackage()` is a constructor that can create a simple container for working with geopackages from several types of inputs. Often you will have a _character_ file path to a GeoPackage (.gpkg) file. 

```{r}
g <- geopackage(gpkg_tmp, connect = TRUE)
g
class(g)
```

Other times you may have a list of tables and layers you want to be in a GeoPackage that does not exist yet. 

```{r}
g2 <- geopackage(list(dem = r, bbox = v))
g2
class(g2)
```

Note that a temporary GeoPackage (`{r, eval=exists(g2)} gpkg_source(g2)`) is automatically created when using the `geopackage(<list>)` constructor. 

You also may have a _DBIConnection_ to a GeoPackage database already opened that you want to use. In any case (_character_, _list_, _SQLiteConnection_) there is an S3 method to facilitate creating the basic _geopackage_ class provided by {gpkg}. All other methods are designed to be able to work smoothly with _geopackage_ class input.

## Inspect Contents of GeoPackage

We can list the table names in a GeoPackage with `gpkg_list_tables()` and fetch pointers (SpatRaster, SpatVectorProxy, and lazy data.frame) to the data in them with `gpkg_table()`. We can check the status of the internal `geopackage` class `SQLiteConnection` with `gpkg_is_connected()` and disconnect it with `gpkg_disconnect()`.

```{r}
# enumerate tables
gpkg_list_tables(g)

# inspect tables
gpkg_tables(g)

# inspect a specific table
gpkg_table(g, "myattr", collect = TRUE)
```

Note that the `collect = TRUE` forces data be loaded into R memory for vector and attribute data; this is the difference in result object class of _SpatVectorProxy_/_SpatVector_ and _tbl_SQLiteConnection_/_data.frame_ for vector and attribute data, respectively.

`gpkg_collect()` is a helper method to call `gpkg_table(..., collect = TRUE)` for in-memory loading of specific tables.

```{r}
gpkg_collect(g, "DEM1")
```

Note that with grid data returned from `gpkg_collect()` you get a table result with the tile contents in a blob column of a _data.frame_ instead of _SpatRaster_ object.

The inverse function of `gpkg_collect()` is `gpkg_tbl()` which always returns a _tbl_SQLiteConnection_.

```{r}
tb <- gpkg_tbl(g, "gpkg_contents")
tb
```

Note that with this lazy reference to table, the internal _SQLiteConnection_ in `g` is the same as the source in `tb`:

```{r}
gpkg_connection(g)@ptr
gpkg_connection(tb)@ptr
```

This means that you can handle all of your connections via the `connect` argument when creating a geopackage object, along with the `gpkg_connect()` and `gpkg_disconnect()` commands.

There are also a variety of convenience functions for the standard required tables in a geopackage. 

For instance, `gpkg_contents()` collects the `"gpkg_contents"` table.

```{r}
gpkg_contents(g)
```

### Lazy Data Access

There are several other methods that can be used for working with tabular data in a GeoPackage in a "lazy" fashion.

#### Method 1: `gpkg_table_pragma()`

`gpkg_table_pragma()` is a low-frills `data.frame` result containing important table information, but not values.  The `PRAGMA table_info()` is stored as a nested data.frame `table_info`. This representation has no dependencies beyond {RSQLite} and is efficient for inspection of table structure and attributes, though it is less useful for data analysis.

```{r}
head(gpkg_table_pragma(g))
```

#### Method 2:  `gpkg_vect()` and `gpkg_query()`

`gpkg_vect()` is a wrapper around `terra::vect()` you can use to create 'terra' `SpatVector` objects from the tables found in a GeoPackage. 
```{r}
gpkg_vect(g, 'bbox')
```

The table of interest need not have a geometry column, but this method does not work on GeoPackage that contain only gridded data, and some layer in the GeoPackage must have some geometry. 

```{r}
gpkg_vect(g, 'gpkg_ogr_contents')
```

The _SpatVectorProxy_ is used for "lazy" references to of vector and attribute contents of a GeoPackage; this object for vector data is analogous to the _SpatRaster_ for gridded data. The 'terra' package provides "GDAL plumbing" for filter and query utilities. 

`gpkg_query()` by default uses the 'RSQLite' driver, but the richer capabilities of OGR data sources can be harnessed with [SQLite SQL dialect](https://gdal.org/en/stable/user/sql_sqlite_dialect.html). These additional features can be utilized with the `ogr=TRUE` argument to `gpkg_query()`, or `gpkg_ogr_query()` for short. This assumes that GDAL is built with support for SQLite (and ideally also with support for Spatialite).

For example, we use built-in functions such as `ST_MinX()` to calculate summaries for `"bbox"` table, geometry column `"geom"`. In this case we expect the calculated quantities to match the coordinates/boundaries of the bounding box:

```{r}
res <- gpkg_ogr_query(g, "SELECT 
                           ST_MinX(geom) AS xmin,
                           ST_MinY(geom) AS ymin, 
                           ST_MaxX(geom) AS xmax, 
                           ST_MaxY(geom) AS ymax 
                          FROM bbox")
as.data.frame(res)
```

#### Method 3: `gpkg_rast()`

Using `gpkg_rast()` you can quickly access references to all tile/gridded datasets in a GeoPackage.

For example: 

```{r}
gpkg_rast(g)
```

#### Method 4: `gpkg_table()`

With the `gpkg_table()` method you access a specific table (by name) and get a "lazy" `tibble` object referencing that table. 

This is achieved via {dplyr} and the {dbplyr} database connection to the GeoPackage via the {RSQLite} driver. The resulting object's data can be used in more complex analyses by using other {dbplyr}/{tidyverse} functions.

For example, we inspect the contents of the `gpkg_contents` table that contains critical information on the data contained in a GeoPackage.

```{r}
gpkg_table(g, "gpkg_contents")
```

As a more complicated example we access the `gpkg_2d_gridded_tile_ancillary` table, and perform some data processing.

We `dplyr::select()` `mean` and `std_dev` columns from the `dplyr::filter()`ed rows where `tpudt_name == "DEM2"`. Finally we materialize a `tibble`  with `dplyr::collect()`:

```{r}
library(dplyr, warn.conflicts = FALSE)

gpkg_table(g, "gpkg_2d_gridded_tile_ancillary") %>% 
  filter(tpudt_name == "DEM2") %>% 
  select(mean, std_dev) %>% 
  collect()
```

### Managing Connections

Several helper methods are available for checking GeoPackage `SQLiteConnection` status, as well as connecting and disconnecting an existing `geopackage` object (`g`).

```{r}
# still connected
gpkg_is_connected(g)

# disconnect 
gpkg_disconnect(g)

# reconnect
gpkg_connect(g)

# disconnect
gpkg_disconnect(g)
```
